Imputation of incomplete multilevel data

### 2024-06-06 manuscript.qmd
draft for writing course

---
title: "Imputation of incomplete multilevel data"
author: "Hanne Oberman and Gerko Vink"
format:
  html:
    toc: true
---

@@

```{r}
library(knitr)
# print everything as kable ----
knit_print.data.frame <- function (x, options, ...) {
  knitr::kable(x) |> knitr::knit_print(options, ...)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame)
options(scipen = 999, digits = 3)
```

# Introduction
This is a tutorial on multiple imputation of incomplete multilevel data with [mice]{.pkg} in [R]{.proglang}. But before we dive into the details and feel sorry for all the holes in our data, it is important to go a few steps back. Because it is not just holes in our data. Most often, there is more data than holes. And more data means more information. In this tutorial we frame your mind to consider the flow of that information in such a way that the multilevel structure is taken into account. In other words, we will show you how to solve for the incomplete data to fit your multilevel model.

@@

## Intended audience
This tutorial is aimed at researchers and analysts who know their way around multilevel analyses, but lack the skills and expertise of dealing with incomplete sources of information. For them we will gently introduce, explain and demonstrate the necessary methodology and its application in [mice]{.pkg}. For applied researchers and analysts that are both unfamiliar with multilevel modeling and incomplete data analysis, we will try to be as complete as possible, but we may refer to other sources to complement our tutorial.

# Dealing with unavailable information
First and foremost, the best way to deal with unavailable information is to make sure that you have none. This is in itself impossible to verify, as one would need the potentially unavailable information to definitively prove that there is no deviation from the conclusions obtained on the available information. Since this is impossible to do in practice, it is far more convenient to assume some model that relates the observed and unobserved parts in terms of the problem at hand. For such models, there are generally three scenarios that we need to consider.

## Scenario 1: Missingness is independent of any information
The first scenario is that analysis of the available information by itself yields the correct conclusion. This is an ideal scenario, but it is most likely not the case. Rubin (1987) defines this scenario as **missing completely at random** (MCAR). Hand (2020) defines this - perhaps more intuitively - as *not data dependent*. There is, however, a caveat with us using Hand's terminology, as the term *data dependent* may lull the reader into the false sense of security of assuming that data refers to the available data. While Hand is careful in making this distinction in his book, we would like to explicitly highlight the possibility that more data is always available, and that not considering this data may lead to incorrect conclusions. We therefore favor the explicit use of the term *missingness* in Rubin's terminology and like to stick with Rubin's MCAR in our text.

@@

## Scenario 2: Missingness depends on available information
This assumption is violated when the probability of being unobserved - or observed, for that matter - is dependent on more than one dimension in the available information. For example when the probability of being unobserved is dependent on the value of another variable in the data, we have a clear and proven deviation from MCAR. This would bring us to the second scenario: **missing at random (MAR)** (Rubin, 1976). With MAR the observed information holds the key to unlocking the missing information. In Hand's terminology, this would be *seen data dependent*. The difficulty here is that - unlike with MCAR, where the missingness mechanism may not be dependent on any observed nor unobserved data - with MAR, it may only be observed data dependent.

@@

## Scenario 3: Missingness may also relate to unavailable information
But what if your missingness is related to the data that you cannot see? This would mean that no matter what type of analysis you perform on the observed information, your conclusions would be invalid. This is the third scenario: **missing not at random (MNAR)** (Rubin, 1976). Hand (2020) defines this as *unseen data dependent*. This is the most difficult scenario to deal with, as it is impossible to model the missing information based on the observed information alone and some form of *adjustment* is needed to arrive at the correct conclusion.

::: {.callout-note}
# Distinguishing beteen missingness mechanisms. 
Everyone who learns about missingness will at some point try to distinguish between the three mechanisms. This is a good exercise, but it is important to note that the distinction is not always clear nor possible.

@@

@@

Setup `R` environment
```{r, message=FALSE, warning=FALSE}
library(mice)         # for imputation
library(miceadds)     # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(broom)        # for tidying model output
library(broom.mixed)  # for tidying model output
```

@@

@@

```{r}
mod_true_singlelevel <- lm(popularity_ij ~ 1, data = popular) 
```

```{r}
mod_true_singlelevel |> tidy()
```

```{r}
mod_true_intercept <- lmer(popularity_ij ~ (1 | cluster_id), data = popular, REML = FALSE) 
```

```{r}
mod_true_intercept |> tidy()
```

```{r}
mod_true_predictors <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id), data = popular, REML = FALSE)
```

```{r}
mod_true_predictors |> tidy()
```

```{r}
mod_true_randomslope <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1  + extraversion_ij | cluster_id), data = popular, REML = FALSE)
```

```{r}
mod_true_randomslope |> tidy()
```

```{r}
mod_true_interaction <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id), data = popular, REML = FALSE)
```

```{r}
mod_true_interaction |> tidy()
```

@@

```{r}
mod_cca_singlelevel <- lm(popularity_ij ~ 1, data = dat) 
```

```{r}
mod_cca_singlelevel |> tidy()
```

```{r}
mod_cca_intercept <- lmer(popularity_ij ~ (1 | cluster_id), data = dat, REML = FALSE) 
```

```{r}
mod_cca_intercept |> tidy()
```

```{r}
mod_cca_predictors <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id), data = dat, REML = FALSE)
```

```{r}
mod_cca_predictors |> tidy()
```

```{r}
mod_cca_randomslope <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1  + extraversion_ij | cluster_id), data = dat, REML = FALSE)
```

```{r}
mod_cca_randomslope |> tidy()
```

```{r}
mod_cca_interaction <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id), data = dat, REML = FALSE)
```

```{r}
mod_cca_interaction |> tidy()
```

@@

The outcome variable 'popularity' is approximately normally distributed.

@@

A density plot of the incomplete outcome variable 'popularity' shows that aggregated across clusters, the distribution is approximately normal. Within clusters, there is more variability.

@@

We will use `2l.norm`: a multilevel method using the linear mixed model with heterogeneous error variances.

```{r}
meth["popularity_ij"] <- "2l.norm"
```

@@

We can use `2l.pmm` to impute the missing values.

```{r}
meth["extraversion_ij"] <- "2l.pmm"
```

@@

```{r}
ggmice(dat, aes(assessment_ij)) + 
  geom_histogram(fill = "white")
```
We will use `2l.pmm` for this variable too.

@@

#### Single-level model

```{r, echo=TRUE}
mod_cca_singlelevel <- lm(
  popularity_ij ~ 1, 
  data = dat
  )
tidy(cca)
```

```{r, echo=TRUE}
mod_imp_singlelevel <- with(
  imp,
  lm(popularity_ij ~ 1, data = dat)
  ) |> 
  pool()
tidy(fit)
```

#### Intercept-only model
```{r}
cca <- lmer(
  popularity_ij ~ (1 | cluster_id),
  data = dat,
  REML = FALSE
  )
tidy(cca)
```

```{r, echo=TRUE}
fit <- with(
  imp,
  lmer(
    popularity_ij ~ (1 | cluster_id), 
    data = dat, 
    REML = FALSE)
  ) 
pool(fit) |> tidy()
```

```{r, echo=TRUE}
mitml::testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

#### Model with explanatory variables

```{r}
cca <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id),
  data = dat,
  REML = FALSE
  )
tidy(cca)
```

```{r, echo=TRUE}
fit <- with(
  imp,
  lmer(
    popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id),
    data = dat, 
    REML = FALSE)
  ) 
pool(fit) |> tidy()
```

```{r, echo=TRUE}
mitml::testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

#### Model with explanatory variables, extraversion slope random

```{r}
cca <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j +
    (1  + extraversion_ij | cluster_id),
  data = dat,
  REML = FALSE
  )
tidy(cca)
```

```{r, echo=TRUE}
fit <- with(
  imp,
  lmer(
    popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
      (1  + extraversion_ij | cluster_id), 
    data = dat, 
    REML = FALSE)
  ) 
pool(fit) |> tidy()
```

```{r, echo=TRUE}
mitml::testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

```{r}
cca <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j +
    extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id),
  data = dat,
  REML = FALSE
  )
tidy(cca)
```

```{r, echo=TRUE}
fit <- with(
  imp,
  lmer(
    popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
      extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id),  
    data = dat, 
    REML = FALSE)
  ) 
pool(fit) |> tidy()
```

```{r, echo=TRUE}
mitml::testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

>>>>

---
title: "Imputation of incomplete multilevel data"
author: "Hanne Oberman and Gerko Vink"
format:
  html:
    toc: true
    self-contained: true
    number-sections: true
---

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
# print everything as kable ----
knit_print.data.frame <- function (x, options, ...) {
  knitr::kable(x) |> knitr::knit_print(options, ...)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame)
options(scipen = 999, digits = 3)
```

# Introduction

Multilevel data are clustered data structures, for example with students clustered in classes or patients clustered in hospitals. Such hierarchical data structures may be accommodated using multilevel modeling---a statistical technique to capture associations in the data both at the unit-level (level-1) as well as the cluster-level (level-2; REF: Hox). Multilevel modeling is a step-by-step procedure (cf. Hox et al., 2017). First, a null-model is fitted to serve as a benchmark. Then, level-1 effects, level-2 effects and cross-level interactions can be added sequentially. The procedure of multilevel modeling can be complicated, especially in the presence of missing values.

If there are missing entries in multilevel data, even the null-model cannot be estimated. That’s why statistical software often defaults to ignoring any incomplete rows in the data. This method, known as ‘list-wise deletion’, allows the analyst to estimate statistical models on the subset of complete cases in the data. Ignoring incomplete cases assumes that the complete cases are representative of the entire sample. If there is a difference between complete and incomplete cases, complete-case analysis will yield biased results (REF: van Buuren). Moreover, with multilevel data structures, missingness can occur at the unit-level (i.e., incomplete cases), but also at the cluster-level (i.e., incomplete or entirely unobserved cluster-level variables). If the observations for a variable are systematically missing in a certain cluster, none of the units in this cluster can be included in the analysis of scientific interest after list-wise deletion. Complete-case analysis then is a wasteful way to accommodate missing data, and may even further bias the analysis results. Hence, the default method to handle missingness in multilevel data is not appropriate.

A statistically valid way to handle incomplete data is to impute (i.e., fill in) every missing data entry multiple times. With multiple imputation, several completed versions of the incomplete data are created, which can be analyzed as if the data were complete. The resulting statistics can then be pooled according to Rubin’s rules (REF: Rubin). Since the analysis results may vary across imputations, pooled estimates will reflect the uncertainty due to non-response. Multiple imputation has been established as a valid all-round method to deal with incomplete data.

This tutorial focuses on one ‘flavor’ of multiple imputation: multiple imputation by chained equations (MICE) as implemented in the R package {mice}. MICE is an algorithmic approach to fill in missing data entries on a variable-by-variable basis. That means that for every incomplete variable in the data, an imputation model should be chosen. This imputation model should be congenial (i.e., statistically matching) the analysis model one is intending to fit after treating the missingness. There is no encompassing imputation model (‘joint model’) required. The flexibility that this approach gives, may also be a disadvantage. If it is not clear at the imputation stage what the final analysis model will be at the multilevel modeling stage, it may be difficult to select appropriate imputation models. Each imputation model should therefore be compatible with the broadest multilevel model the analyst is intending to fit after imputation. In this tutorial, we will outline how to build imputation models that are in line with the multilevel structure in the data.

The aim of this tutorial is to provide empirical researchers who are faced with incomplete multilevel data guidance in imputing the missing values in their data. Missing data pose a wicked, but treatable problem. Combined with multilevel structures (another wicked but treatable problem), the solutions are a lot more difficult. This tutorial will aid empirical researchers in validly handling incomplete multilevel data by means of multiple imputation with the R package {mice}. Other valid methods (such as smcfcs) are outside the scope of this tutorial. No experience with missing data and imputation is required. We will assume basic familiarity with multilevel modeling. The notation will follow Hox et al. (2017). The R package {lme4} is used for the analysis of scientific interest. Analysis estimates after imputation are pooled with {miceadds}. Visualizations for data exploration and evaluation of the imputations will be performed using {ggplot2}  and {ggmice}.

The requirements to follow along with this tutorial are an incomplete dataset with a multilevel structure, and the conviction that there are no reasons for the missingness that cannot be modeled from the data (ignorability assumption, see Box XYZ). Software-wise, this tutorial requires R and the packages mice, lme4, ggplot2, ggmice and miceadds. As a case study, we will use an adapted version of the popularity data, published by Hox et al. (2017). The original dataset does not contain any missing values. For the purpose of this tutorial, we have created an ‘amputed’ (i.e. incomplete) version of the data. The complete version will serve as comparative truth in this tutorial. Both the complete and incomplete data can be downloaded from XYZ. By the end of this tutorial, the reader will be able to build an imputation model for each incomplete variable, evaluate the imputation models, impute the data, evaluate the imputations, and analyze the data, and evaluate the effects of the imputations.

::: {.callout-note}
# Ignorability
Missingness is canonically categorized into three missing data mechanisms. The terminology proposed by Rubin (1987). In this paper we will assume MCAR or MAR. MCAR means that there is no relation between the missingness and observed values. MAR means that there are dependencies between the missingness and the data, but these dependencies can be modeled from the data. The case of MNAR is outside of the scope of this tutorial. We will not cover missingness that is related to unseen parts of the data, e.g. the reason for the missingness depends on the missing value itself.
:::

# Theoretical background [OLD text, do not review]

Before we dive into the details and feel sorry for all the holes in our data, it is important to go a few steps back. Because it is not just holes in our data. Most often, there is more data than holes. And more data means more information. In this tutorial we frame your mind to consider the flow of that information in such a way that the multilevel structure is taken into account. In other words, we will show you how to solve for the incomplete data to fit your multilevel model.

@@

This tutorial is aimed at researchers and analysts who know their way around multilevel analyses, but lack the skills and expertise of dealing with incomplete sources of information. For them we will gently introduce, explain and demonstrate the necessary methodology and its application in [mice]{.pkg}. For applied researchers and analysts that are both unfamiliar with multilevel modeling and incomplete data analysis, we will try to be as complete as possible, but we may refer to other sources to complement our tutorial.

## Dealing with unavailable information
First and foremost, the best way to deal with unavailable information is to make sure that you have none. This is in itself impossible to verify, as one would need the potentially unavailable information to definitively prove that there is no deviation from the conclusions obtained on the available information. Since this is impossible to do in practice, it is far more convenient to assume some model that relates the observed and unobserved parts in terms of the problem at hand. For such models, there are generally three scenarios that we need to consider.

### Scenario 1: Missingness is independent of any information
The first scenario is that analysis of the available information by itself yields the correct conclusion. This is an ideal scenario, but it is most likely not the case. Rubin (1987) defines this scenario as **missing completely at random** (MCAR). Hand (2020) defines this - perhaps more intuitively - as *not data dependent*. There is, however, a caveat with us using Hand's terminology, as the term *data dependent* may lull the reader into the false sense of security of assuming that data refers to the available data. While Hand is careful in making this distinction in his book, we would like to explicitly highlight the possibility that more data is always available, and that not considering this data may lead to incorrect conclusions. We therefore favor the explicit use of the term *missingness* in Rubin's terminology and like to stick with Rubin's MCAR in our text.

@@

### Scenario 2: Missingness depends on available information
This assumption is violated when the probability of being unobserved - or observed, for that matter - is dependent on more than one dimension in the available information. For example when the probability of being unobserved is dependent on the value of another variable in the data, we have a clear and proven deviation from MCAR. This would bring us to the second scenario: **missing at random (MAR)** (Rubin, 1976). With MAR the observed information holds the key to unlocking the missing information. In Hand's terminology, this would be *seen data dependent*. The difficulty here is that - unlike with MCAR, where the missingness mechanism may not be dependent on any observed nor unobserved data - with MAR, it may only be observed data dependent.

@@

### Scenario 3: Missingness may also relate to unavailable information
But what if your missingness is related to the data that you cannot see? This would mean that no matter what type of analysis you perform on the observed information, your conclusions would be invalid. This is the third scenario: **missing not at random (MNAR)** (Rubin, 1976). Hand (2020) defines this as *unseen data dependent*. This is the most difficult scenario to deal with, as it is impossible to model the missing information based on the observed information alone and some form of *adjustment* is needed to arrive at the correct conclusion.

::: {.callout-note}
# Distinguishing beteen missingness mechanisms 
Everyone who learns about missingness will at some point try to distinguish between the three mechanisms. This is a good exercise, but it is important to note that the distinction is not always clear nor possible.

@@

In this tutorial, we will use the `R` package {`mice`} for multiple imputation by chained equations, {`lme4`} for multilevel analysis, {`miceadds`} to obtain estimates of multilevel effects after imputation, and {`ggplot2`} and {`ggmice`} for data visualization.

The {`mice`} package provides tools for multivariate imputation by chained equations. The main function, `mice()`, can be run on any dataset in tabular form. To fill in the missing values, the function requires an incomplete dataset. All other arguments, such as the number of imputations `m`, are optional. 
The `mice()`  functions has sensible defaults. Some of these defaults, however, are not applicable to multilevel data. Following the recommendations by van Buuren (2018), each incomplete variable should be imputed using an appropriate imputation model. An imputation model is a combination of an imputation method (the functional form of the model) and relevant imputation model predictors (other variables in the data that are associated with the incomplete variable and/or the missingness indicator of the incomplete variable). In the context of multilevel data, the functional form of the imputation models will likely need a “2l” component. The “2l” methods take the multilevel structure of the data into account. The `method` argument in the `mice()` function can be used to specify multilevel imputation methods for every incomplete variable in the data. Then the `predictorMatrix` argument can be used to supply imputation model predictors. The clustering variable is denoted with the value `-2`, unit-level associations (or level-1 effects) are denoted with `2`, and cluster-level associations (level-2 effects) are denoted with `1`.

The {`lme4`} package allows for multilevel modeling. The `lmer()` function requires a formula to define the multilevel model and a dataset to estimate the model on. After fitting multilevel models, the {miceadds} package can be used to obtain pooled estimates of multilevel effects.

The ‘tidyverse’ package {`ggplot2`} is a data visualization package. The `ggplot()` function takes a dataset and mapping instructions. With the family of `geom_*()` functions, different plots can be made. The {ggmice} package extends the functionality of {`ggplot2`}. The `ggmice()` function is equivalent to `ggplot()` except from the way it handles missing or imputed data. The {`ggmice`} package also contains specialized functions for the visualization of e.g. imputation models.

Together, these packages provide functionality treat, analyze, and evaluate incomplete multilevel data. Tabel XYZ gives an overview of the steps that we deem necessary in an imputation workflow for multilevel data. Please note that this list is not exhaustive; we propose these steps as a minimum requirement, accompanied by our `R` function suggestions.

1.	Inspect incomplete data
  a.	Explore variables: determine imputation methods with str() and ggmice(), 
  b.	If necessary, add variables for polynomial terms
  c.	Explore missingness: evaluate missing data patterns with plot_pattern()
  d.	Determine imputation model predictors using plot_corr(), plot_flux(), and ggmice()
2.	Build imputation models
  a.	Create methods vector with imputation methods using make.methods()
  b.	Create predictor matrix with imputation model predictors using make.predictorMatrix()
  c.	Evaluate imputation models using plot_pred()
3.	Run imputation algorithm
  a.	Impute: fill in the missing values with mice()
  b.	Inspect convergence: visualize algorithmic convergence with plot_trace()
  c.	If necessary, add iterations
4.	Evaluate imputations
  a.	Inspect imputed values: assess imputed values with ggmice(), PPC?
  b.	Analyze imputed data: fit the model of scientific interest with lmer()  
  c.	Evaluate missingness metrics with testEstimates()
  d.	If necessary, add imputations
5.	Interpret results and report

## Imputation with `mice` [OLD text, do not review]

@@

Setup `R` environment
```{r, message=FALSE, warning=FALSE}
library(mice)         # for imputation
library(miceadds)     # for imputation
library(micemd)       # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(broom)        # for tidying model output
library(broom.mixed)  # for tidying model output
```

@@

Load the data, make sure the variables are correctly formatted (e.g. numeric clustering variable)

@@

```{r}
mod_true_singlelevel <- lm(
  popularity_ij ~ 1, 
  data = popular
  ) 
```

```{r}
mod_true_singlelevel |> tidy(conf.int = TRUE)
```

```{r}
mod_true_intercept <- lmer(
  popularity_ij ~ (1 | cluster_id), 
  data = popular, 
  REML = FALSE
  ) 
```

```{r}
mod_true_intercept |> tidy(conf.int = TRUE)
```

```{r}
mod_true_predictors <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id), 
  data = popular, 
  REML = FALSE
  )
```

```{r}
mod_true_predictors |> tidy(conf.int = TRUE)
```

```{r}
mod_true_randomslope <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
    (1  + extraversion_ij | cluster_id), 
  data = popular, 
  REML = FALSE
  )
```

```{r}
mod_true_randomslope |> tidy(conf.int = TRUE)
```

```{r}
mod_true_interaction <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
    extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id), 
  data = popular, 
  REML = FALSE
  )
```

```{r}
mod_true_interaction |> tidy(conf.int = TRUE)
```

@@

```{r}
mod_cca_singlelevel <- lm(
  popularity_ij ~ 1, 
  data = dat
  ) 
```

```{r}
mod_cca_singlelevel |> tidy(conf.int = TRUE)
```

```{r}
mod_cca_intercept <- lmer(
  popularity_ij ~ (1 | cluster_id), 
  data = dat, 
  REML = FALSE
  ) 
```

```{r}
mod_cca_intercept |> tidy(conf.int = TRUE)
```

```{r}
mod_cca_predictors <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
    (1 | cluster_id), 
  data = dat, 
  REML = FALSE
  )
```

```{r}
mod_cca_predictors |> tidy(conf.int = TRUE)
```

```{r}
mod_cca_randomslope <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
    (1  + extraversion_ij | cluster_id), 
  data = dat, 
  REML = FALSE
  )
```

```{r}
mod_cca_randomslope |> tidy(conf.int = TRUE)
```

```{r}
mod_cca_interaction <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j +
    extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id),
  data = dat,
  REML = FALSE
  )
```

```{r}
mod_cca_interaction |> tidy(conf.int = TRUE)
```

@@

The outcome variable 'popularity' is approximately normally distributed aggregated over clusters.

@@

This density plot of the incomplete outcome variable 'popularity' shows that aggregated across clusters, the distribution is approximately normal. Within clusters, however, there is more variability.

@@

The variable is not normally distributed within clusters. We will use `2l.pmm`: a semi-parametric multilevel method using the linear mixed model with homogeneous error variances.

```{r}
meth["popularity_ij"] <- "2l.pmm"
```

@@

We will use `2l.pmm` for this variable too.
```{r}
meth["extraversion_ij"] <- "2l.pmm"
```

@@

```{r}
ggmice(dat, aes(assessment_ij)) + 
  geom_histogram(fill = "white")
```

We will use `2l.pmm` to impute the missing values.
<!-- , which is similar to `2l.2stage.pmm` but with homogeneous error variances. -->

@@

Single-level model.
```{r, echo=TRUE}
mod_imp_singlelevel <- with(
  imp,
  lm(popularity_ij ~ 1)
  ) 
```

```{r, echo=TRUE}
pool(mod_imp_singlelevel) |> tidy(conf.int = TRUE)
```

Intercept-only model.
```{r}
mod_imp_intercept <- with(
  imp,
  lmer(
    popularity_ij ~ (1 | cluster_id), 
    REML = FALSE)
  ) 
```

```{r}
pool(mod_imp_intercept) |> tidy(conf.int = TRUE)
```

```{r, echo=TRUE}
mitml::testEstimates(as.mitml.result(mod_imp_intercept), extra.pars = TRUE)
```

Model with explanatory variables.

```{r}
mod_imp_predictors <- with(
  imp,
  lmer(
    popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id),
    REML = FALSE)
  )
```

```{r}
pool(mod_imp_predictors) |> tidy(conf.int = TRUE)
```

```{r, echo=TRUE}
mitml::testEstimates(as.mitml.result(mod_imp_predictors), extra.pars = TRUE)
```

Model with explanatory variables, extraversion slope random.

```{r}
mod_imp_randomslope <- with(
  imp,
  lmer(
    popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
      (1  + extraversion_ij | cluster_id))
  ) 
```

Analysis model does not converge.

```{r}
mod_imp_randomslope <- with(
  imp,
  lmer(
    popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
      (1  + extraversion_ij | cluster_id), 
    control = lmerControl(optimizer = "nloptwrap", optCtrl = list(maxfun = 2e5))
    )
  ) 
```

A different optimizer or extra iterations do not help.

```{r}
pool(mod_imp_randomslope) |> tidy(conf.int = TRUE)
```

```{r, echo=TRUE}
mitml::testEstimates(as.mitml.result(mod_imp_randomslope), extra.pars = TRUE)
```

@@

```{r}
mod_imp_interaction <- with(
  imp,
  lmer(
    popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
      extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id),  
    REML = FALSE, 
    control = lmerControl(optimizer = "bobyqa"))
  ) 
```

```{r}
pool(mod_imp_interaction) |> tidy(conf.int = TRUE)
```

```{r, echo=TRUE}
mitml::testEstimates(as.mitml.result(mod_imp_interaction), extra.pars = TRUE)
```

# Discussion

This tutorial has outlined how to treat missingness in multilevel data using multiple imputation. Multiple imputation with the R package {mice} is a powerful tool to obtain valid estimates of multilevel model effects from incomplete data. Although the procedure of imputation is less effortless than the default missing data method---list-wise deletion---imputation will lead to less biased estimates and confidence-valid inferences. Moreover, data visualization may aid the analyst in building and evaluating imputation models. Since there is software available to facilitate the exploration-imputation-evaluation-analysis pipeline, it is worth the investment. Sometimes a sub-optimal solution has to be preferred, e.g. because of convergence issues. Still, doing something with the missingness is better than ignoring the incomplete cases altogether. In short, using multiple imputation to treat incomplete multilevel data is valid and easy with the right software tools.
